---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# lmmise

<!-- badges: start -->
[![R-CMD-check](https://github.com/BmBaczkowski/lmmise/workflows/R-CMD-check/badge.svg)](https://github.com/BmBaczkowski/lmmise/actions/workflows/R-CMD-check.yaml)

[![lint](https://github.com/BmBaczkowski/lmmise/actions/workflows/lint.yaml/badge.svg)](https://github.com/BmBaczkowski/lmmise/actions/workflows/lint.yaml)
<!-- badges: end -->

`lmmise` provides a domain-specific language (DSL) for specifying linear mixed-effects models and exporting them to fitting and simulation environments.

## Core Concepts

- **`lmm_spec()`** — Define the model skeleton (response variable, distribution, link).
- **`set_fixed()`** — Add fixed effects via a formula (e.g., `y ~ 1 + x1 + x2`).
- **`set_random()`** — Add random effects per grouping variable (e.g., `id ~ 1 + x1`). Supports correlated or independent random effects in the fitted model via `model_corr`.
- **`lmm_params()`** — Create a parameter object initialized from a spec.
- **`patch()`** — Fill parameter values (fixed effects, dispersion, random SDs, correlations) into the params object.
- **`as_lme4()`** — Translate the spec to `lme4::lmer()` / `lme4::glmer()` formula and family.
- **`as_simstudy()`** — Translate the spec + params to `simstudy` arguments for simulating random effects (`type = "rand_effects"`) or outcome data (`type = "outcome"`).
- **`as_lmList()`** — Translate to `lme4::lmList()` formula syntax.

## Installation

You can install the development version of `lmmise` from GitHub with:

``` r
# install.packages("devtools")
devtools::install_github("BmBaczkowski/lmmise")
```

Or using the `remotes` package:

``` r
# install.packages("remotes")
remotes::install_github("BmBaczkowski/lmmise")
```


## Quick Start

### 1. Define the model structure

```{r}
library(lmmise)

# Specify a linear mixed model
spec <- lmm_spec(response = "y", dist = "gaussian") |>
  set_fixed(y ~ 1 + x1 + x2) |>
  set_random(id ~ 1 + x1, model_corr = TRUE) |>
  set_random(site ~ 1)
```

### 2. Set parameter values

```{r}
params <- lmm_params(spec) |>
  patch(list(
    beta = list(intercept = 1.0, x1 = 0.5, x2 = -0.3),
    dispersion = 0.5,
    random_sd = list(
      id = list(intercept = 1.0, x1 = 0.3),
      site = list(intercept = 0.5)
    ),
    random_corr = list(
      id = list(structure = "cs", r = 0.3)
    )
  ))

# confirm all parameters are set
check_params(params)
```

### 3. Export for fitting with lme4

```{r}
lme4_args <- as_lme4(spec, include_random = TRUE)
# lme4_args$formula  -> y ~ 1 + x1 + x2 + (1 + x1 | id) + (1 | site)
# lme4_args$family   -> gaussian(link = "identity")

# Then fit with lme4:
# fit <- lme4::lmer(lme4_args$formula, data = my_data)
```

### 4. Simulate data with simstudy

```{r}
# Random effects for simstudy::addCorData()
re_args <- as_simstudy(spec, params, type = "rand_effects")

# Outcome definition for simstudy::defDataAdd()
out_args <- as_simstudy(spec, params, type = "outcome")
```

## Supported Distributions

- `gaussian` (link: identity)
- `binomial` (link: logit; dispersion fixed at 1)
- `poisson`  (link: log; dispersion fixed at 1)
- `gamma`    (link: inverse)

## Random Effects Syntax

- `id ~ 1`                     — random intercept only
- `id ~ 1 + x1`                — random intercept + slope for `x1`
- `id ~ 1 + x1, model_corr = TRUE`   — correlated effects in the fitted **lme4** model (`(1 + x1 | id)`)
- `id ~ 1 + x1, model_corr = FALSE`  — uncorrelated effects in the fitted **lme4** model (`(1 | id) + (0 + x1 | id)`)

The `model_corr` flag controls how random effects are written to the **lme4** formula.  It does *not* affect simulation correlation.  Simulation correlation is set separately in `patch(random_corr = ...)` using per-group entries.  Supported structures for simulation are:

- **Independent (IND)** — `list(structure = "ind")` (diagonal correlation).
- **Compound symmetry (CS)** — `list(structure = "cs", r = 0.3)` produces an exchangeable correlation with off-diagonal `r`.
- **AR(1)** — `list(structure = "ar1", r = 0.5)` produces an AR(1) matrix where correlation decays as `r^|i-j|` according to term order.
- **Toeplitz** — `list(structure = "toep", r = c(0.5, 0.2))` for a banded Toeplitz; supply `n-1` lags for `n` random terms.
- **Unstructured (UN)** — `list(matrix = M)` where `M` is a square correlation matrix whose row/column names match the random terms (e.g., `"intercept"`, `"x1"`).


## License

MIT + file LICENSE
